---
title: "Massachusetts_Analysis"
author: "Emma Wellbaum"
date: "4/22/2021"
output: html_document
editor_options:
  chunk_output_type: console
---
## Set up

```{r setup, include=FALSE}
getwd()

#install.packages("auk")
library(auk)
library(tidyr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(viridis)
library(cowplot)
library(corrplot)

#Set theme
mytheme <- theme_bw(base_size = 14) + 
  theme(axis.text = element_text(color = "black"),
         legend.position = "bottom")
theme_set(mytheme)
```

## Import and Summarize State Temperature Data

```{r temp.data.analysis}
# Import state temperature data, downloaded from the National Atmospheric and
# Oceanic Administration using the rnoaa package
MA_temp <- read.csv('./Data/Raw/Massachusetts/MassTemperature20102021.csv') %>% unique()

# Summarize monthly temperature data for the entire state
MA_temp <- MA_temp %>%
  mutate(Month_Year = paste0(Month,'-',Year)) %>% 
  mutate(Month_Year = my(Month_Year)) %>%
  drop_na(AvgMonthlyTemp) %>% 
  group_by(Month_Year) %>% 
  summarize(AvgMonthlyTemp_F = mean(AvgMonthlyTemp)) %>%
  filter(year(Month_Year) %in% c(2010:2020))
```

## Import Bird Observation Data

```{r import.bird.data, include=FALSE}
###Import bird observation data downloaded from eBird
#Osprey
OSPR <- read_ebd("./Data/Raw/Massachusetts/ebd_US-MA_osprey_relFeb-2021/ebd_US-MA_osprey_relFeb-2021.txt")

#Red-winged Blackbird 
#Due to large file size, RWBL data was imported in two parts and then combined
RWBL_early <- read_ebd("./Data/Raw/Massachusetts/ebd_US-MA_rewbla_201001_201412_relMar-2021/ebd_US-MA_rewbla_201001_201412_relMar-2021.txt")
RWBL_late <- read_ebd("./Data/Raw/Massachusetts/ebd_US-MA_rewbla_201501_202012_relMar-2021/ebd_US-MA_rewbla_201501_202012_relMar-2021.txt")

#Combine the two RWBL dataframes into a dataframe for 2010-2020
RWBL <- bind_rows(RWBL_early, RWBL_late)

#Wood Duck
WODU <- read_ebd(
  "./Data/Raw/Massachusetts/ebd_US-MA_wooduc_relFeb-2021/ebd_US-MA_wooduc_relFeb-2021.txt")

#Create a merged dataset with all three species
MAbirds <- bind_rows(OSPR, RWBL, WODU)
```

## Clean and Filter Bird Data

```{r clean.bird.data, include=FALSE}
###Clean and filter the merged bird observation data
MAbirds <- MAbirds %>% 
  #Filter species for columns of interest
  select(common_name:observation_count, state, county, latitude:observation_date, protocol_type, duration_minutes) %>% 
  #Create/adjust columns useful for summarizing data
  mutate(#Replace "presence" notations ('X') in observation_count with values of 1
         observation_count = as.numeric(replace(
           observation_count, observation_count == 'X', '1')),
         #Add an observations_per_min column to control for increased effort
         observations_per_min = observation_count/duration_minutes,
         #Add a binary "Presence" column
         Presence = 1,
         #Add a date column set to the first day of each month
         Month_Year = floor_date(observation_date, unit = 'month')) %>% 
  #Filter for years 2010-2020 (when there is data for all three species)
  filter(year(Month_Year) %in% c(2010:2020))


#Exclude NA and Inf values in the observations_per_min column
MAbirds <- MAbirds %>% 
  filter( is.na(observations_per_min) == FALSE) %>% 
  filter( observations_per_min != Inf)
#This filtering removed 5.1% of the original three-species dataset (21,781 rows).

#Examine the results
summary(MAbirds$observations_per_min)
#Visualize distribution
ggplot(MAbirds, aes(x=observations_per_min)) + 
  geom_density() 
ggplot(MAbirds, aes(x=log10(observations_per_min))) + 
  geom_density() 
#The normal distribution did not show up on a plot, but the log10 transformed data looks decent


#### Create summary dataset of bird data by month ####
# Statistics per species: (1) sum of observation_count per month; (2) sum of presence per month; and (3) average observations per minute per month
MAbirds_month <- MAbirds %>% 
  group_by(common_name, Month_Year) %>% 
  summarize(state = first(state),
            observations_per_min_avg = mean(observations_per_min),
            presence_count = sum(Presence),
            observation_count = sum(observation_count)) %>%
  select(common_name, state, observation_count,  observations_per_min_avg, presence_count, Month_Year)

##Join the bird observation to the temperature data
MAbirds_temp <- full_join(MAbirds_month, MA_temp, by='Month_Year')


##Create numeric Month and Year columns for correlation plots
MAbirds_temp <- MAbirds_temp %>% 
  mutate(Month = month(Month_Year),
         Year = year(Month_Year))

#Create a  numeric "Season" column based on Month and the meteorological seasons:
#Winter=1 (Dec-Feb), Spring=2 (Mar=May), Summer=3 (Jun-Aug), Fall=4 (Sept-Nov)
#The value mapping is easiest to remember by singing "You've Got a Friend in Me"
MAbirds_temp$Season[MAbirds_temp$Month %in% c(1,2,12)] <- 1 #Winter
MAbirds_temp$Season[MAbirds_temp$Month %in% c(3:5)] <- 2 #Spring
MAbirds_temp$Season[MAbirds_temp$Month %in% c(6:8)] <- 3 #Summer
MAbirds_temp$Season[MAbirds_temp$Month %in% c(9:11)] <- 4 #Fall

#Save output to csv
write.csv(MAbirds_temp, row.names = FALSE, './Data/Processed/Massachusetts/MAbirds_temp.csv')
```

## Exploratory Plots

```{r monthly.sum.plots}
#Create a basic density plot for all species across the entire date range
#This plot represents bird observations uncorrected for effort
ggplot(MAbirds, aes(x = observation_date, fill = common_name, color = common_name)) +
  geom_density(alpha = .2) + 
  facet_wrap(~common_name, ncol = 3) + 
  theme(legend.position = 'none',
         axis.text.x = element_text(angle = 45, hjust = 1)) + 
  #Set color and fill
  scale_color_brewer( palette = 'Dark2') + 
  scale_fill_brewer( palette = 'Dark2') + 
  #Create figure labels
  labs(x = "Observation Date", title="Bird Observation Density by Species",
       subtitle = "Massachusetts: 2010:2020")


#Plot Average Observations per Minute for 2010-2021 by Species
ggplot(MAbirds_temp, 
       aes(x=Month_Year, y=log10(observations_per_min_avg), color=AvgMonthlyTemp_F)) +
  geom_line(lwd=0.8) +
  #Set color to average monthly temperature
  scale_color_viridis(option='plasma') +
  #Create legible date break labels
  scale_x_date(date_breaks = '1 year', date_labels = '%Y') +
  #Create figure labels
  labs(x='Observation Date', 
       y='Average Observations/Minute (log10)', 
       color='Mean Monthly Temperature (F)', 
       title= 'Effort-Corrected Bird Observations by Species', 
       subtitle = 'Massachusetts: 2010-2020') +
  #Facet by species
  facet_wrap(~common_name, nrow=3)

#### Cowplot of average observations per minute by year and species ####

#Osprey component plot (controls the y-axis label)
OSPRplot <- ggplot(filter(MAbirds_temp, common_name == 'Osprey'), 
                   aes(x=month(Month_Year), y=log10(observations_per_min_avg),
                       group=year(Month_Year), color=AvgMonthlyTemp_F)) +
  geom_line(lwd=0.8) +
  #Set color to average monthly temperature
  scale_color_viridis(option='plasma') + 
  #scale x and y axes the same for each component plot
  scale_y_continuous(limits = c(-1.75, 0.5), n.breaks = 3) +
  scale_x_continuous(expand = c(0,0), breaks = c(1, 4, 8, 12), 
                     labels = c('Jan','Apr', 'Aug', 'Dec')) +
  labs(x='', y='Average Observations/Minute (log10)', color='', title = 'Osprey') +
  #Create legible figure labels (for the combined cowplot)
  theme(legend.position='none', title=element_text(size=10), 
        axis.text.y=element_text(size=9), axis.title.y= element_text(size=10),
        strip.text=element_text(size=9)) +
  #Facet by year
  facet_grid(vars(year(Month_Year)))


#Red-winged Blackbird component plot (controls the legend)
RWBLplot <- ggplot(filter(MAbirds_temp, common_name == 'Red-winged Blackbird'),
                   aes(x=month(Month_Year), y=log10(observations_per_min_avg),
                       group=year(Month_Year), color=AvgMonthlyTemp_F)) +
  geom_line(lwd=0.8) +
  scale_color_viridis(option='plasma') + 
  scale_y_continuous(limits = c(-1.75, 0.5), n.breaks = 3) +
  scale_x_continuous(expand = c(0,0), breaks = c(1, 4, 8, 12), 
                     labels = c('Jan','Apr', 'Aug', 'Dec')) +
  labs(x='', y='', color='Average Monthly Temperature (F)', 
       title='Red-winged Blackbird') +
  theme(legend.position='bottom', title=element_text(size=10),
        axis.text.y=element_text(size=9), strip.text=element_text(size=9)) +
  facet_grid(vars(year(Month_Year)))


#Wood Duck component plot
WODUplot <- ggplot(filter(MAbirds_temp, common_name == 'Wood Duck'),
                          aes(x=month(Month_Year), y=log10(observations_per_min_avg),
                              group=year(Month_Year), color=AvgMonthlyTemp_F)) +
  geom_line(lwd=0.8) +
  scale_color_viridis(option='plasma') + 
  scale_y_continuous(limits = c(-1.75, 0.5), n.breaks = 3) +
  scale_x_continuous(expand = c(0,0), breaks = c(1, 4, 8, 12), 
                     labels = c('Jan','Apr', 'Aug', 'Dec')) +
  labs(x='', y='', color='', title='Wood Duck') +
  theme(legend.position='none', title=element_text(size=10), 
        axis.text.y=element_text(size=9), strip.text=element_text(size=9)) +
  facet_grid(vars(year(Month_Year)))


#Extract the legend from the RWBL component plot as an object
birdleg <-get_legend(RWBLplot)

#Create a title for the combined plot
title <- ggdraw() + 
  draw_label('Seasonal Stay in Massachusetts by Species: 2010:2020',
             fontface = 'bold',
             x = 0,
             hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 5))

#Combine the individual species plots into 3-species cowplot
birdplot <- plot_grid(OSPRplot, RWBLplot + theme(legend.position='none'), WODUplot, align = c('hv'), nrow=1)
                        
#Add the title and legend to birdplot 
birdplot <- plot_grid(title, birdplot, birdleg, ncol = 1, rel_heights = c(.08, 1.2, .15))
print(birdplot)
```


## Statistical Analysis

> Question: For each bird species, what is the relationship between the average monthly observations per minute and average monthly temperature?

> Is month (i.e., time) or average monthly temperature a better predictor of species-specific trends in average monthly observations per minute?

> Hypothesis: This is likely species depended. I expect that temperature will have the largest affect on the  warmer temperatures will result in less birds, effect of temperature will depend on month.

```{r subset}
#Subset data by species
OSPR_MA_temp <- MAbirds_temp %>% filter(common_name == "Osprey")
RWBL_MA_temp <- MAbirds_temp %>% filter(common_name == "Red-winged Blackbird")
WODU_MA_temp <- MAbirds_temp %>% filter(common_name == "Wood Duck")
```

## Correlation Plots
> What does the relationship between bird obersations, time, and temperature look like?
> Which numeric date variables would be most interesting to explore with respect to changes in temperature and bird observations?
> How are changes in temperature associated with time in Massachusetts?

```{r correlation.tests}
## Question: How are changes in temperature associated with time in Massachusetts?
cor.test(MAbirds_temp$Month, MAbirds_temp$AvgMonthlyTemp_F)
cor.test(MAbirds_temp$Season, MAbirds_temp$AvgMonthlyTemp_F)
cor.test(MAbirds_temp$Year, MAbirds_temp$AvgMonthlyTemp_F)
```
> Answer: Temperature is not associated with year (cor= -0.01547243). This indicates that we would need to download data for a significantly larger date range to observe potential species shifts related to climate change. There are stronger associations between temperature and both month and season (cor=0.2572472 and 0.5924802, respectively). This is expected given the known seasonal variation in Massachusetts.

```{r correlation.plots}
# Create correlation plots of the numeric variables in our processed dataset: 
# observation count, presence, average observations per minute, month, year, temp, and season

## Osprey
corMatrix_OSPR <- cor(OSPR_MA_temp[,c(3:5,7:10)])
corrplot(corMatrix_OSPR, method = "color")
#Strongest relationship between month and the observational variables, which makes sense because the Osprey is the species that leaves Massachusetts entirely of the three observed

## Red winged Blackbird 
corMatrix_RWBL <- cor(RWBL_MA_temp[,c(3:5,7:10)])
corrplot(corMatrix_RWBL, method = "color")
# Strong negative correlation between observations per minute and average monthly temperature

## Wood Duck
corMatrix_WODU <- cor(WODU_MA_temp[,c(3:5,7:10)])
corrplot(corMatrix_WODU, method = "color")
#Least exciting plot
```
> The most evident, non-direct associations from the correlation plots are those between observations per minute and average monthly temperature, observations per mintue and month, and observations per minute and season. Across the three species, there are also potential relationships between temperautre and time variables and observation count and/or presence. However, the statistical analysis for MA focuses on the effort-adjusted observation data.


## Linear Regressions
> Question 1: For each bird species, what is the relationship between the average monthly observations per minute and average monthly temperature?

> Question 2: Is month, season, (i.e., time), or average monthly temperature a better predictor of species-specific trends in average monthly observations per minute?

> Hypothesis: This is likely species depended. I expect that temperature will have the largest affect on the  warmer temperatures will result in less birds, effect of temperature will depend on month.

```{r linear.regressions}

#### Osprey ####
#Observations per minute vs. Temperature and Month
lm_OSPR_month <- lm(data = OSPR_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + 
                as.factor(Month))
#Summarize output
summary(lm_OSPR_month)


#Observations per minute vs. Temperature and Season
lm_OSPR_season <- lm(data = OSPR_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + 
                as.factor(Season))
#Summarize output
summary(lm_OSPR_season)

#### Red winged Blackbird ####
#Observations per minute vs. Temperature and Month
lm_RWBL_month <- lm(data = RWBL_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + 
                as.factor(Month))
#Summarize output
summary(lm_RWBL_month)


#Observations per minute vs. Temperature and Season
lm_RWBL_season <- lm(data = RWBL_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + 
                as.factor(Season))
#Summarize output
summary(lm_RWBL_season)


#### Wood Duck ####
#Observations per minute vs. Temperature and Month
lm_WODU_month <- lm(data = WODU_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + 
                as.factor(Month))
#Summarize output
summary(lm_WODU_month)

#Observations per minute vs. Temperature and Season
lm_WODU_season <- lm(data = WODU_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + 
                as.factor(Season))
#Summarize output
summary(lm_WODU_season)

```
> Analysis: