---
title: "Massachusetts_Analysis"
author: "Emma Wellbaum"
date: "4/22/2021"
output: pdf_document
editor_options:
  chunk_output_type: console
---

## Set up

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# set working directory for knitting
knitr::opts_knit$set( root.dir = 
  #"/Users/Kate/Documents/1.Spring 2021/JaffeWellbaumFrear_ENV872_EDA_FinalProject",
    "/Users/emmaw/Documents/ENV872/JaffeWellbaumFrear_ENV872_EDA_FinalProject",
                  tidy.opts = list(width.cutoff = 60),
                  tidy = TRUE)
getwd()

#install.packages("auk")
library(auk)
library(tidyr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(viridis)
library(cowplot)
library(corrplot)
library(agricolae)

#Set theme
mytheme <- theme_bw(base_size = 14) + 
  theme(axis.text = element_text(color = "black"),
         legend.position = "bottom")
theme_set(mytheme)
```

## Import and Summarize State Temperature Data
> Temperature data was obtained from the National Centers for Environmental Information Global Historical Climate Network Daily dataset. The GHCN-Daily contains daily climate summaries, including minimum and maximum temperature, from land surface stations across the world. Data was downloaded via the National Atmospheric and Oceanic Administration API using the rnoaa package and a NOAA API token. The source code used to extract and pre-process the data is located in the project repository file: './Code/TemperatureAPI.Rmd'

```{r temp.data.analysis}
# Import state temperature data, downloaded from the National Atmospheric and
# Oceanic Administration using the rnoaa package
# The source code used to download and pre-process the temperature data is
MA_temp <- read.csv('./Data/Raw/Massachusetts/MassTemperature20102021.csv') %>% unique()

# Summarize monthly temperature data for the entire state
MA_temp <- MA_temp %>%
  mutate(Month_Year = paste0(Month,'-',Year)) %>% 
  mutate(Month_Year = my(Month_Year)) %>%
  drop_na(AvgMonthlyTemp) %>% 
  group_by(Month_Year) %>% 
  summarize(AvgMonthlyTemp_F = mean(AvgMonthlyTemp)) %>%
  filter(year(Month_Year) %in% c(2010:2020))
```

## Import Bird Observation Data
> Bird observation data was downloaded from the Cornell Lab of Ornithology eBird Database (https://ebird.org/).

```{r import.bird.data}
###Import bird observation data downloaded from eBird
#Osprey
OSPR <- read_ebd(
  "./Data/Raw/Massachusetts/ebd_US-MA_osprey_relFeb-2021/ebd_US-MA_osprey_relFeb-2021.txt")

#Red-winged Blackbird 
#Due to large file size, RWBL data was imported in two parts and then combined
RWBL_early <- read_ebd(
  "./Data/Raw/Massachusetts/ebd_US-MA_rewbla_201001_201412_relMar-2021/ebd_US-MA_rewbla_201001_201412_relMar-2021.txt")
RWBL_late <- read_ebd(
  "./Data/Raw/Massachusetts/ebd_US-MA_rewbla_201501_202012_relMar-2021/ebd_US-MA_rewbla_201501_202012_relMar-2021.txt")

#Combine the two RWBL dataframes into a dataframe for 2010-2020
RWBL <- bind_rows(RWBL_early, RWBL_late)

#Wood Duck
WODU <- read_ebd(
  "./Data/Raw/Massachusetts/ebd_US-MA_wooduc_relFeb-2021/ebd_US-MA_wooduc_relFeb-2021.txt")

#Create a merged dataset with all three species
MAbirds <- bind_rows(OSPR, RWBL, WODU)
```

## Clean and Filter Bird Data

```{r clean.bird.data}
###Clean and filter the merged bird observation data
MAbirds <- MAbirds %>% 
  #Filter species for columns of interest
  select(common_name:observation_count, state, county, 
         latitude:observation_date, protocol_type, duration_minutes) %>% 
  #Create/adjust columns useful for summarizing data
  mutate(#Replace "presence" notations ('X') in observation_count with values of 1
         observation_count = as.numeric(replace(
           observation_count, observation_count == 'X', '1')),
         #Add an observations_per_min column to control for increased effort
         observations_per_min = observation_count/duration_minutes,
         #Add a binary "Presence" column
         Presence = 1,
         #Add a date column set to the first day of each month
         Month_Year = floor_date(observation_date, unit = 'month')) %>% 
  #Filter for years 2010-2020 (when there is data for all three species)
  filter(year(Month_Year) %in% c(2010:2020))


#Exclude NA and Inf values in the observations_per_min column
MAbirds <- MAbirds %>% 
  filter( is.na(observations_per_min) == FALSE) %>% 
  filter( observations_per_min != Inf)
#This filtering removed 5.1% of the original three-species dataset (21,781 rows).

#Examine the results
summary(MAbirds$observations_per_min)
#Visualize distribution
ggplot(MAbirds, aes(x=observations_per_min)) + 
  geom_density() #Normal distribution has a heavy right skew
ggplot(MAbirds, aes(x=log10(observations_per_min))) + 
  geom_density() #The log10 transformed data looks decent

#Save output to csv
write.csv(MAbirds, row.names = FALSE, 
          './Data/Processed/Massachusetts/MAbirds.csv')
```

## Summarize MA Bird Data

```{r summarize.bird.data, message=FALSE, warning=FALSE}
## Create summary dataset of bird data by month
#Statistics per species: (1) sum of observation_count per month,
#(2) sum of presence per month, and (3) average observations per minute per month
MAbirds_month <- MAbirds %>% 
  group_by(common_name, Month_Year) %>% 
  summarize(state = first(state),
            observations_per_min_avg = mean(observations_per_min),
            presence_count = sum(Presence),
            observation_count = sum(observation_count)) %>%
  select(common_name, state, observation_count, 
         observations_per_min_avg, presence_count, Month_Year)

##Join the bird observation to the temperature data
MAbirds_temp <- full_join(MAbirds_month, MA_temp, by='Month_Year')


#### Prepare the data for statistical analysis ####

#Create numeric Month and Year columns for correlation plots
MAbirds_temp <- MAbirds_temp %>% 
  mutate(Month = month(Month_Year),
         Year = year(Month_Year))

#Create a  numeric "Season" column based on Month and the meteorological seasons:
#Winter=1 (Dec-Feb), Spring=2 (Mar=May), Summer=3 (Jun-Aug), Fall=4 (Sept-Nov)
#The value mapping is easiest to remember by singing "You've Got a Friend in Me"
MAbirds_temp$Season[MAbirds_temp$Month %in% c(1,2,12)] <- 1 #Winter
MAbirds_temp$Season[MAbirds_temp$Month %in% c(3:5)] <- 2 #Spring
MAbirds_temp$Season[MAbirds_temp$Month %in% c(6:8)] <- 3 #Summer
MAbirds_temp$Season[MAbirds_temp$Month %in% c(9:11)] <- 4 #Fall

#Save output to csv
write.csv(MAbirds_temp, row.names = FALSE, 
          './Data/Processed/Massachusetts/MAbirds_temp.csv')
```

## MA Exploratory Plots

```{r density.plot, fig.cap="Basic density plot for all MA species across the entire date range. This plot represents bird observations uncorrected for effort.", echo=FALSE}
ggplot(MAbirds, aes(x = observation_date, fill = common_name, color = common_name)) +
  geom_density(alpha = .2) + 
  facet_wrap(~common_name, ncol = 3) + 
  theme(legend.position = 'none',
         axis.text.x = element_text(angle = 45, hjust = 1)) + 
  #Set color and fill
  scale_color_brewer( palette = 'Dark2') + 
  scale_fill_brewer( palette = 'Dark2') + 
  #Create figure labels
  labs(x = "Observation Date", title="Bird Observation Density by Species",
       subtitle = "Massachusetts: 2010:2020")
```


```{r monthly.sum.plot, fig.cap="Average Observatons per Minute during 2020-2020 by Species in Massachusetts", fig.height=5, echo=FALSE}
ggplot(MAbirds_temp, 
       aes(x=Month_Year, y=log10(observations_per_min_avg), color=AvgMonthlyTemp_F)) +
  geom_line(lwd=0.8) +
  #Set color to average monthly temperature
  scale_color_viridis(option='plasma') +
  #Create legible date break labels
  scale_x_date(date_breaks = '1 year', date_labels = '%Y') +
  #Create figure labels
  labs(x='Observation Date', 
       y='Avg. Observations/Minute (log10)', 
       color='Mean Monthly Temperature (F)', 
       title= 'Effort-Corrected Bird Observations by Species', 
       subtitle = 'Massachusetts: 2010-2020') +
  #Facet by species
  facet_wrap(~common_name, nrow=3)
```


```{r MA.cowplot, fig.cap="Cowplot of Average Observations per Minute by Year and Species in Massachusetts", echo=FALSE}

#Osprey component plot (controls the y-axis label)
OSPRplot <- ggplot(filter(MAbirds_temp, common_name == 'Osprey'), 
                   aes(x=month(Month_Year), y=log10(observations_per_min_avg),
                       group=year(Month_Year), color=AvgMonthlyTemp_F)) +
  geom_line(lwd=0.8) +
  #Set color to average monthly temperature
  scale_color_viridis(option='plasma') + 
  #scale x and y axes the same for each component plot
  scale_y_continuous(limits = c(-1.75, 0.5), n.breaks = 3) +
  scale_x_continuous(expand = c(0,0), breaks = c(1, 4, 8, 12), 
                     labels = c('Jan','Apr', 'Aug', 'Dec')) +
  labs(x='', y='Average Observations/Minute (log10)', color='', title = 'Osprey') +
  #Create legible figure labels (for the combined cowplot)
  theme(legend.position='none', title=element_text(size=10), 
        axis.text.y=element_text(size=9), axis.title.y= element_text(size=10),
        strip.text=element_text(size=9)) +
  #Facet by year
  facet_grid(vars(year(Month_Year)))


#Red-winged Blackbird component plot (controls the legend)
RWBLplot <- ggplot(filter(MAbirds_temp, common_name == 'Red-winged Blackbird'),
                   aes(x=month(Month_Year), y=log10(observations_per_min_avg),
                       group=year(Month_Year), color=AvgMonthlyTemp_F)) +
  geom_line(lwd=0.8) +
  scale_color_viridis(option='plasma') + 
  scale_y_continuous(limits = c(-1.75, 0.5), n.breaks = 3) +
  scale_x_continuous(expand = c(0,0), breaks = c(1, 4, 8, 12), 
                     labels = c('Jan','Apr', 'Aug', 'Dec')) +
  labs(x='', y='', color='Average Monthly Temperature (F)', 
       title='Red-winged Blackbird') +
  theme(legend.position='bottom', title=element_text(size=10),
        axis.text.y=element_text(size=9), strip.text=element_text(size=9)) +
  facet_grid(vars(year(Month_Year)))


#Wood Duck component plot
WODUplot <- ggplot(filter(MAbirds_temp, common_name == 'Wood Duck'),
                          aes(x=month(Month_Year), y=log10(observations_per_min_avg),
                              group=year(Month_Year), color=AvgMonthlyTemp_F)) +
  geom_line(lwd=0.8) +
  scale_color_viridis(option='plasma') + 
  scale_y_continuous(limits = c(-1.75, 0.5), n.breaks = 3) +
  scale_x_continuous(expand = c(0,0), breaks = c(1, 4, 8, 12), 
                     labels = c('Jan','Apr', 'Aug', 'Dec')) +
  labs(x='', y='', color='', title='Wood Duck') +
  theme(legend.position='none', title=element_text(size=10), 
        axis.text.y=element_text(size=9), strip.text=element_text(size=9)) +
  facet_grid(vars(year(Month_Year)))

##Combine into a single plot using cowplot

#Extract the legend from the RWBL component plot as an object
birdleg <-get_legend(RWBLplot)
#Create a title for the combined plot
title <- ggdraw() + 
  draw_label('Seasonal Stay in Massachusetts by Species: 2010:2020',
             fontface = 'bold',
             x = 0,
             hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 5))
#Combine the individual species plots into 3-species cowplot
birdplot <- plot_grid(OSPRplot, RWBLplot + theme(legend.position='none'),
                      WODUplot, align = c('hv'), nrow=1)
#Add the title and legend to birdplot 
birdplot <- plot_grid(title, birdplot, birdleg, ncol = 1, 
                      rel_heights = c(.08, 1.2, .15))
#Print
print(birdplot)
```

## MA Statistical Analysis

> Question: For each bird species, what is the relationship between the average monthly observations per minute and average monthly temperature?

> Is month (i.e., time) or average monthly temperature a better predictor of species-specific trends in average monthly observations per minute?

> Hypothesis: This is likely species depended. I expect that temperature will have the largest affect on the warmer temperatures will result in less birds, effect of temperature will depend on month.

```{r subset.MA, include=FALSE}
#Subset data by species
OSPR_MA_temp <- MAbirds_temp %>% filter(common_name == "Osprey")
RWBL_MA_temp <- MAbirds_temp %>% filter(common_name == "Red-winged Blackbird")
WODU_MA_temp <- MAbirds_temp %>% filter(common_name == "Wood Duck")

#Save the species subsets to .csv
write.csv(OSPR_MA_temp, row.names = FALSE, 
          './Data/Processed/Massachusetts/OSPR_MA_temp.csv')
write.csv(RWBL_MA_temp, row.names = FALSE, 
          './Data/Processed/Massachusetts/RWBL_MA_temp.csv')
write.csv(WODU_MA_temp, row.names = FALSE, 
          './Data/Processed/Massachusetts/WODU_MA_temp.csv')
```

### Correlation Tests
> Question: How are changes in temperature associated with time in Massachusetts?

```{r correlation.tests}
cor.test(MAbirds_temp$Month, MAbirds_temp$AvgMonthlyTemp_F)
cor.test(MAbirds_temp$Season, MAbirds_temp$AvgMonthlyTemp_F)
cor.test(MAbirds_temp$Year, MAbirds_temp$AvgMonthlyTemp_F)
```
> Answer: Temperature is not associated with year (cor= -0.01547243). This indicates that we would need to download data for a significantly larger date range to observe potential species shifts related to climate change. There are stronger associations between temperature and both month and season (cor=0.2572472 and 0.5924802, respectively). This is expected given the known seasonal variation in Massachusetts.

### Correlation Plots
> Create correlation plots for the numeric variables in our processed dataset: observation count, presence, average observations per minute, month, year, temp, and season
> Question: What does the relationship between bird observations, time of year, and temperature look like?

```{r correlation.plots}
## Osprey
corMatrix_OSPR <- cor(OSPR_MA_temp[,c(3:5,7:10)])
corrplot(corMatrix_OSPR, method = "color")

## Red winged Blackbird 
corMatrix_RWBL <- cor(RWBL_MA_temp[,c(3:5,7:10)])
corrplot(corMatrix_RWBL, method = "color")

## Wood Duck
corMatrix_WODU <- cor(WODU_MA_temp[,c(3:5,7:10)])
corrplot(corMatrix_WODU, method = "color")
```
> Answer: Based on the correlation plots, the strongest, non-colinear relationships appear to be between observations per minute and average monthly temperature, observations per minute and month, and observations per minute and season. The osprey correlation plot had the strongest association between month and the observational variables of the three species while the red-winged blackbird ahd the strongest association between temperature and the observational variables (specifically observatons per minute). Across the three species, there also appear to be associations between temperature/time and observation count/presence. However, the statistical analysis for MA will focus on effort-adjusted observation data only.


### Linear Regressions
> Question: For each species, is month, season, or average monthly temperature a better predictor of effort-adjusted observations (average observations per minute)?

> NOTE: For each species, Month and Season are evaluated separately (with Average Monthly Temperature) to avoid regressions colinearity. Season is a numeric variable that was created by grouping the Month variable into four groups, which will cause singularities when running a regression.

> Hypothesis: Time of year (i.e., month or season) will have a stronger relationship to observations per minute than average monthly temperature. In other words, there will be a stronger association with the time of year that birds tend to be observed than trends in temperature associated with those times of year. Between month and season, I expect that month will be a better predictor of effort-adjusted observations for the osprey, who are not present in Massachusetts year round. I expect the grouping of months into seasons will make the the broader season variable be a better predictor of bird-observations for the red-winged blackbird and wood duck, both of which have resident populations.

```{r linear.regressions}
#### Osprey ####
#Observations per minute vs.Temperature and Month
lm_OSPR.MA.month <- lm(data = OSPR_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + as.factor(Month)-1)
#Summarize output
summary(lm_OSPR.MA.month)
#Select model based on AIC
step(lm_OSPR.MA.month) #no recommended change to model


#Observations per minute vs. Temperature and Season
lm_OSPR.MA.season <- lm(data = OSPR_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + as.factor(Season)-1)
#Summarize output
summary(lm_OSPR.MA.season)
#Select model based on AIC
step(lm_OSPR.MA.season) #no recommended change to model

#### Red winged Blackbird ####

#Observations per minute vs. Temperature and Month 
lm_RWBL.MA.month <- lm(data = RWBL_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + as.factor(Month)-1)
#Summarize output
summary(lm_RWBL.MA.month)
#Select model based on AIC
step(lm_RWBL.MA.month) #stepwise function recommends dropping temperature

##Observations per minute vs. Month 
lm_RWBL.MA.month.only <- lm(data = RWBL_MA_temp,
                observations_per_min_avg ~ as.factor(Month)-1)
#Summarize output
summary(lm_RWBL.MA.month.only)
step(lm_RWBL.MA.month.only)


#Observations per minute vs. Temperature and Season 
lm_RWBL.MA.season <- lm(data = RWBL_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + as.factor(Season)-1)
#Summarize output
summary(lm_RWBL.MA.season)
#Select model based on AIC
step(lm_RWBL.MA.season) #no recommended change to model


#### Wood Duck ####

#Observations per minute vs. Temperature and Month 
lm_WODU.MA.month <- lm(data = WODU_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + as.factor(Month)-1)
#Summarize output
summary(lm_WODU.MA.month)
#Select model based on AIC
step(lm_WODU.MA.month) #no recommended change to model


#Observations per minute vs. Temperature and Season
lm_WODU.MA.season <- lm(data = WODU_MA_temp,
                observations_per_min_avg ~ AvgMonthlyTemp_F + as.factor(Season)-1)
#Summarize output
summary(lm_WODU.MA.season)
#Select model based on AIC
step(lm_WODU.MA.season) #no recommended change to model

```

### Analysis 

> **Osprey:** The osprey month-temperature model explained slightly more variance in effort-adjusted observations than the season-temperature model(**84.23%** and **84.05%**, respectively). Looking at the month and season coefficients, this is likely due to the lack/absence of osprey observational data during January and February (assumed to be from migration). The months surrounding this period (i.e., March, April, November, December) are the most statistically significant in the month-temperature model (p-value <0.001). Given that three seasons are represented in those four months, it makes sense that modeling the data by month was more accurate than modeling the data by season for this species. Average monthly temperature was statistically significant in both models: month-temperature p-value: **0.027345** and season-temperature p-value: **0.000945**. This result was expected, as the correlation between season and temperature was greater than the correlation between month and temperature based on earlier correlation tests. That being said, the coefficient of average monthly temperature was not more significant than any of the seasonal coefficients for the osprey, which supports the hypothesis that time of year is a better predictor of changes in bird observations than temperature.

> **Red-winged Blackbird:** Similar to the osprey, the red-winged blackbird month-temperature model explained the most variance in effort-corrected observations -- **74.28%**. However, unlike the osprey, the stepwise function for the red-winged blackbird month-temperature model recommended removing average monthly temperature from the model (**Starting AIC: -302.77; Final AIC: -304.24**). Prior to removing temperature, the red-winged blackbird month-temperature model explained 74.16% of the variance in observation data, which was still a better fit than the season-temperature model (explaining 71.74% of variance). This result is consistent with the hypothesis that time of year is a better predictor of bird observation effort than temperature. While only ten months had statistically significant p-values > 0.5, month is still appears to be a better predictor of changes in effort-corrected observations for the red-winged blackbird than season. Seasonal trends in observations may be difficult to analyze for the red-winged blackbird in Massachusetts given the prevalence of the species and the fact that there are both resident and migratory populations. Looking at the month and season factors together, the change that occurs in species observations during the winter months is definitely worth exploring (p-value for Season 1 (Winter) is the most significant: 2.79e-15). However, I would recommend attempting to narrow down observations to a subpopulation of birds in Massachusetts.

> **Wood Duck:** The wood duck month-temperature model explained more variance in effort-adjusted observations than the season-temperature model by the largest margin of the three bird species -- **87.77%** to **83.98%**. Similar to the osprey, average monthly temperature was statistically significant for both models, but had a larger impact on bird observations in the season-temperature model (season-temperature p-value: 0.00239; month-temperature p-value: 0.026586). The stepwise functions for the wood duck models did not recommend dropping any variables or factors. Like the red-winged blackbird, wood ducks have both a resident and migratory population in Massachusetts. The resident population of wood ducks can make it difficult to evaluate their migration patterns using observation data because the species is observed year round. Viewing the month and season regressions together, there appear to be differences in wood duck observations in the Spring & Summer vs. Fall & Winter, which could be interesting to explore in more detail.


### Post-hoc Analysis
> For each species, what months have similar effort-adjusted bird observations?
> While not explored in great detail here, these groupings may indicate a more focused period of time to study migratory patterns in the future.

```{r post.hoc.analysis}
#### Osprey ####

#Fit an observations per minute vs. month model as an anova
aov_OSPR.MA.month <- 
  aov(data = OSPR_MA_temp, observations_per_min_avg ~ as.factor(Month))
#Summarize the aov
summary(aov_OSPR.MA.month)

#Reveal groupings with similar bird observations: Tukey HSD Test
OSPR.MA.month.grps <- 
  HSD.test(aov_OSPR.MA.month, "as.factor(Month)", group = TRUE)
#Print the groupings
OSPR.MA.month.grps


#### Red-winged Blackbird ####

#Fit an observations per minute vs. month model as an anova
aov_RWBL.MA.month <- 
  aov(data = RWBL_MA_temp, observations_per_min_avg ~ as.factor(Month))
#Summarize the aov
summary(aov_OSPR.MA.month)

# Reveal groupings with similar bird observations: Tukey HSD Test
RWBL.MA.month.grps <- 
  HSD.test(aov_RWBL.MA.month, "as.factor(Month)", group = TRUE)
#Print the groupings
RWBL.MA.month.grps


#### Wood Duck ####

#Fit an observations per minute vs. month model as an anova
aov_WODU.MA.month <- 
  aov(data = WODU_MA_temp, observations_per_min_avg ~ as.factor(Month))
#Summarize the aov
summary(aov_WODU.MA.month)

#Reveal groupings with similar bird observations: Tukey HSD Test
WODU.MA.month.grps <- 
  HSD.test(aov_WODU.MA.month, "as.factor(Month)", group = TRUE)
#Print the groupings
WODU.MA.month.grps
```

