---
title: "North Carolina Final Analysis"
author: "Cate Jaffe"
date: "4/21/2021"
output: pdf_document
editor_options:
  chunk_output_type: console
---

### Setup
```{r setup}

##### Library #####
library(tidyverse)
library(ggplot2)
library(scales)
library(auk) # eBird Package
library(agricolae) 
library(lubridate)
library(corrplot)
library(colormap)
library(ggridges)
library(RColorBrewer)

# set working directory for knitting
knitr::opts_knit$set( root.dir = 
                        "/Users/Kate/Documents/1.Spring 2021/JaffeWellbaumFrear_ENV872_EDA_FinalProject",
                  tidy.opts = list(width.cutoff = 60),
                  tidy = TRUE)
# check wd
getwd()

# set ggplot theme
mytheme <- theme_light( base_size = 14) + 
  theme( axis.text = element_text( color = "#222222ff"),
         legend.position = "top",
         # margins (top,right,bottom,left)
         axis.title.x = element_text( color = "black",
                                    margin = margin(20,0,0,0)),
         axis.title.y = element_text( color = "black",
                                    margin = margin(0,20,0,0)))
theme_set(mytheme)
```

### Import Data
```{r data import}
# import bird data
# eBird data is in text format, package "auk" used to convert to dataframe
woodduck <- read_ebd("./Data/Raw/NorthCarolina/ebd_US-NC_wooduc_relFeb-2021/ebd_US-NC_wooduc_relFeb-2021.txt")
rwbbird <- read_ebd("./Data/Raw/NorthCarolina/ebd_US-NC_rewbla_relMar-2021/ebd_US-NC_rewbla_relMar-2021.txt")
osprey <- read_ebd("./Data/Raw/NorthCarolina/ebd_US-NC_osprey_relFeb-2021/ebd_US-NC_osprey_relFeb-2021.txt")

# import temperature data
temp <- read.csv("./Data/Raw/NorthCarolina/NCTemperature20102021.csv") %>% unique()

```

### Data Cleaning
```{r bird data clean}

### create merged dataset of all bird data
allBirds <- bind_rows(woodduck, rwbbird, osprey)

### clean
allBirds <- allBirds %>%
  # filter dates to date range of interest: 2010 - 2021
  filter(year(observation_date) > 2009 & year(observation_date) < 2022) %>%
  # select only columns of interest
  select(common_name:observation_count, state, county,
                                      latitude:time_observations_started, protocol_type,
                                      duration_minutes:number_observers,
                                      all_species_reported) %>% 
    # change "X" value in observation_count to 1 (X represents "present" in eBird)
    # change to numeric variable after converting to X
    mutate(observation_count = as.numeric(replace(observation_count, observation_count == 'X', '1')),
           # add column which divides # observations per minute observation
           # this controls for birding "effort" which was much higher in later years (2015-2020)
           observations_per_min = observation_count/duration_minutes,
           # add column for "Year-Month" using floor_date from lubridate
           Year_Month = floor_date(observation_date, unit = "month"),
           # add binary presence column
           Presence = 1) %>% 
  # Some "observations per minute" values are NA or Inf where the duration of observation was 0 minutes or was missing. Exclude NA and Inf values
  filter( is.na(observations_per_min) == FALSE) %>% 
  filter( observations_per_min != Inf)
  # Removing NA and Inf values removes about 10,000 observations, or ~6% of the data.

### check distribution of new "observations per minute" variable
summary(allBirds$observations_per_min)
# plot regular distribution:
# heavy right skew in data
ggplot(allBirds, aes(x=observations_per_min)) + 
  geom_density() 
# log transformation looks more normal
ggplot(allBirds, aes(x=log10(observations_per_min))) + 
  geom_density() 


### Create grouped dataset, with observations per minute averaged by year-month
allBirds_YMgrouped <- 
  allBirds %>%
  group_by(common_name, Year_Month) %>% 
  summarize(state = first(state),
            observation_count_sum = sum(observation_count),
            # sum presence column: how many times did bird appear on a checklist?
            presence_count = sum(Presence),
            # average obsv/min: how many times was the bird observed, corrected for effort
            observation_per_min_avg = mean(observations_per_min)
            ) 

```

```{r temperature data cleaning}

# group by month & year
temp_YM <- temp %>% 
  # omit NAs
  na.omit() %>% 
  # make Year-Month column
  mutate( Year_Month = ydm((paste0(Year,"-01-",Month)))) %>% 
  group_by(Year_Month) %>% 
  # take statewide average
  summarise(AvgMonthlyTemp_Statewide = mean(AvgMonthlyTemp)) %>% 
  # re-add month and year columns
  mutate(Month = month(Year_Month),
         Year = year(Year_Month))

  
```

### Data Exploration
```{r bird exploratory plots}

### Birds, ungrouped, uncorrected for effort
# basic density plot across all years, faceted 
ggplot(allBirds, aes(x = observation_date, fill = common_name, color = common_name)) +
  geom_density(alpha = .2) + 
  facet_wrap(vars(common_name), nrow = 1, ncol = 3) + 
  theme( legend.position = "none",
         axis.text.x = element_text(angle = 45,
                                    hjust = 1)) + 
  # disable scientific notation in y axis
  # set color and fill manually
  scale_color_brewer( palette = "Dark2") + 
  scale_fill_brewer( palette = "Dark2") + 
  # make labels legible
  labs(x = "Observation Date")

### Birds, grouped by year-month, corrected for effort
ggplot(allBirds_YMgrouped %>% filter(year(Year_Month) != 2021),
       aes(x = Year_Month, y = log10(observation_per_min_avg), group = common_name, color = common_name)) +
  geom_line(alpha = .8) + 
  facet_wrap(vars(common_name), nrow = 1, ncol = 3) + 
  theme( legend.position = "none",
         axis.text.x = element_text(angle = 45,
                                    hjust = 1)) + 
  # disable scientific notation in y axis
  # set color and fill manually
  scale_color_brewer( palette = "Dark2") + 
  scale_fill_brewer( palette = "Dark2") + 
  # make labels legible
  labs(x = "Observation Date")

```

```{r temperature exploratory plots}

# Temperature trends by year-month groups
ggplot(temp_YM %>% filter(Year != 2021), aes(x = Year_Month, y = AvgMonthlyTemp_Statewide, color = AvgMonthlyTemp_Statewide)) +
  geom_point(alpha = .4) +
  geom_line(alpha = .8) + 
  scale_x_date(breaks = "6 months",
                date_labels = "%b %Y") + 
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1),
        legend.position = "none") + 
  scale_color_colormap(colormap = "plasma") + 
  labs(y = "Average Monthly Temperature (F)",
       x = "Date")
  

```

### Join Bird and Temperature Data

```{r join temp and bird data}

birdsTemp_YM <- full_join(allBirds_YMgrouped, temp_YM, 
                             by = "Year_Month")

```

### Analysis: Bird Observations & Temperature

```{r Plots}

#  observations per minute to overtime, by temperature
ggplot(birdsTemp_YM %>% filter(Year != 2021), 
       aes(x = Year_Month, y = log10(observation_per_min_avg),
           color = AvgMonthlyTemp_Statewide)) + 
  geom_line(lwd = .8) + 
  facet_wrap(vars(common_name), nrow = 3, scale = "free") + 
  theme(legend.position = "right") + 
  scale_color_colormap(colormap = "plasma") + 
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  labs(y = "Average Birds per Minute of Observation (log10)",
       x = "Year",
       color = "Average Monthly \nTemperature (F)",
       title = "Bird Observations in North Carolina",
       subtitle = "2010 - 2020")


```

# Correlation Plots

```{r corr plots}

## Osprey
ospreyOnly <- birdsTemp_YM %>%
# filter to bird of interest                     
  filter(common_name == "Osprey")

# exclude non-numeric columns with [,4:9]
corMatrix_osprey <- cor(ospreyOnly[,4:9])
corrplot(corMatrix_osprey, method = "color")


## Red winged Blackbird 
rwbbOnly <- birdsTemp_YM %>%
  filter(common_name == "Red-winged Blackbird")

corMatrix_rwbb <- cor(rwbbOnly[,4:9])
corrplot(corMatrix_rwbb, method = "color")


## Wood Duck
woodduckOnly <- birdsTemp_YM %>%
  filter(common_name == "Wood Duck")

corMatrix_woodduck <- cor(woodduckOnly[,4:9])
corrplot(corMatrix_woodduck, method = "color")

```

```{r Statistical Analyses - Linear Regressions}

### Osprey ###########
# Observations per minute vs. Temperature, Year, and Month
# as an lm()
lm_osprey <- lm(data = birdsTemp_YM %>% filter(common_name == "Osprey"),
                observation_per_min_avg ~ AvgMonthlyTemp_Statewide + 
                  # have to convert month and year to factors 
                  as.factor(Year) + as.factor(Month))
# summarize output
summary(lm_osprey)
# stepwise selection of most parsimonious model
step(lm_osprey)

# as an aov()
aov_osprey <- aov(data = birdsTemp_YM %>% filter(common_name == "Osprey"),
                observation_per_min_avg ~ 
                  as.factor(Year) + as.factor(Month))

summary(aov_osprey)

# post analysis Tukey Test, only run with categorical explanatory variables.
# this post analysis test will reveal which groups of years and/or 
# months had similar observations of birds
TukeyHSD(aov_osprey)

# create and print group labels - for Month
osprey.groups.yr <- 
  HSD.test(aov_osprey, "as.factor(Year)", group = TRUE)
osprey.groups.yr

# create and print group labels - for Year
osprey.groups.month <- 
  HSD.test(aov_osprey, "as.factor(Month)", group = TRUE)
osprey.groups.month


### Red winged Blackbird ###########
lm_rwbb <- lm(data = birdsTemp_YM %>% filter(common_name == "Red-winged Blackbird"),
                observation_per_min_avg ~ AvgMonthlyTemp_Statewide + as.factor(Year) + as.factor(Month))

summary(lm_rwbb)

step(lm_rwbb)

# stepwise selection suggests a model with only month is the most parsimonious 
lm_rwbb_monthOnly <- lm(data = birdsTemp_YM %>% filter(common_name == "Red-winged Blackbird"),
                observation_per_min_avg ~ as.factor(Month))

summary(lm_rwbb_monthOnly)

# run as AOV
aov_rwbb <- aov(data = birdsTemp_YM %>% filter(common_name == "Red-winged Blackbird"),
                observation_per_min_avg ~ as.factor(Month))

summary(aov_rwbb)

# post analysis Tukey Test, only run with categorical explanatory variables.
# this post analysis test will reveal which groups of years and/or 
# months had similar observations of birds
TukeyHSD(aov_rwbb)

# create and print group labels - for Month
rwbb.groups.month <- 
  HSD.test(aov_osprey, "as.factor(Month)", group = TRUE)
rwbb.groups.month

### Wood Duck ###########
lm_duck <- lm(data = birdsTemp_YM %>% filter(common_name == "Wood Duck"),
                observation_per_min_avg ~ AvgMonthlyTemp_Statewide + Year + Month)

summary(lm_duck)

step(lm_duck)
# stepwise selection indicates no variables should be removed from model
# No Tukey HSD is run for this model because it includes a continuous numerical variable (Temperature)


```

> Draft Written Summary/Notes about Linear regressions:

> Osprey: The most parisimonious model for the Osprey observations (corrected for effort) included year and month as explanatory variables, but not temperature. This model explained *68%* of the variation in Osprey observations. Like the Blackbird, the spring and summar months had similar observations (Group ab: April, May, June, August) which were statistically different than the mean of observations in fall and winter months (Group cd: September, October, November, December). 

> Red-Winged Blackbird: The most parisimonious model for the Blackbird observations (corrected for effort) included only month as an explanatory variable. This model explained only 18.9% of the variation in Blackbird observations. Like the Osprey, the spring and summar months had similar observations of Blackbird (Group ab: April, May, June, August) which were statistically different than the mean of observations in fall and winter months (Group cd: September, October, November, December). 

> Wood Duck: The most parisimonious model for the Wood Duck observations (corrected for effort) included temperature, year and month as explanatory variables. Together, these variables explain only 5.6% of the variation in wood duck observations. For every 1 degree *increase* in temperature (with month and year held constant) we would expect the observations of wood ducks (per minute of observation) to *decrease* by .00072 duck per minute observation. There is likely some other variable, not measured here, explaining the variation in wood duck observations in North Carolina between 2010 and 2020.

> Overall, the Wood Duck appears to be the only bird of the three examined in this study for which average monthly temperature has a statistically significant relationship with bird abundance (corrected for observation effort). The Month of the Year was included in the final model for all three birds, and observations tended to be most similar in non-migratory periods (namely late spring to summer and late fall to winter).

### Analysis: Date of First and Last Annual Observations

```{r Visualize first and last observation dates}

allBirdsTemp <- full_join(allBirds, temp_YM, 
                             by = "Year_Month")

#Summarize the number of observations by week
allBirds_week <- allBirdsTemp %>%
  # make column of "week"
  mutate(Week = floor_date(observation_date, unit = "week")) %>% 
  group_by(Week, Year_Month, common_name, state) %>% 
  summarize(observations_per_min_avg = mean(observations_per_min),
            Presence = sum(Presence),
            observation_count_sum = sum(observation_count),
            AvgWeeklyTemp_F = mean(AvgMonthlyTemp_Statewide)) %>%
  mutate(Month = month(Year_Month),
         Year = year(Year_Month)) %>% 
  filter(Year %in% c(2010:2020)) 

# Plot
ggplot(allBirds_week, aes(x=month(Week), y=log10(observation_count_sum), color=common_name)) +
  geom_smooth(method='loess', se=F, lwd=0.8) +
  scale_color_manual(values=c('#1E88E5', '#D81B60', '#004D40')) + 
#  scale_y_continuous(n.breaks = 3) +
  scale_x_continuous(expand = c(0,0), breaks = c(1, 4, 8, 12), labels = c('Jan','Apr', 'Aug', 'Dec')) +
labs(x='', y='', color='', title='Seasonal Stay by Species Over Time', subtitle = 'North Carolina') +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.spacing.x = unit(2, 'lines')) +
  facet_grid(vars(year(Year_Month)), vars(common_name))


```

### Discussion of Seasonal Stays by each species
> All birds appear to reside in North Carolina year round. Osprey appear most populus in the summer months and almost absent in the winter, with populations potentially increasing overtime between 2010 and 2020. The opposite pattern appears to be true for Blackbird, which is more abundant in the summers and less in the winters, and is generally the most populus of the three birds examined here. Wood Duck appears from this plot to be least seaonsal, and to be present in constant numbers (at State scale) throughout the year. 

> To further investigate seasonal and long-term (2010-2020) trends, we conducted a time-series analysis:

### Time Series Analysis

```{r Time Series}

birdsTemp_YM_ospr <- birdsTemp_YM %>% filter(common_name == "Osprey" & Year %in% c(2010:2020))

birdsTemp_YM_rwbb <- birdsTemp_YM %>% filter(common_name == "Red-winged Blackbird" & Year %in% c(2010:2020))

birdsTemp_YM_wodo <- birdsTemp_YM %>% filter(common_name == "Wood Duck" & Year %in% c(2010:2020))


#Osprey
osprey.ts <- ts(OSPR_ME_month$observation_per_min_avg, 
                             start = c(2010,1), frequency = 12)

OSPR.ts.decomposed <- stl(OSPR.ts, s.window = "periodic")
# Visualize the decomposed series. 
plot(OSPR.ts.decomposed)

monthly_OSPR_trend <- Kendall::SeasonalMannKendall(OSPR.ts)
summary(monthly_OSPR_trend)

OSPR.ts.components <- as.data.frame(OSPR.ts.decomposed$time.series[,1:3])


#Red-winged Blackbird

RWBL.ts <- ts(RWBL_ME_month$observation_per_min_avg, 
                             start = c(2010,1), frequency = 12)

RWBL.ts.decomposed <- stl(RWBL.ts, s.window = "periodic")
# Visualize the decomposed series. 
plot(RWBL.ts.decomposed)

monthly_RWBL_trend <- Kendall::SeasonalMannKendall(RWBL.ts)
summary(monthly_RWBL_trend)

RWBL.ts.components <- as.data.frame(RWBL.ts.decomposed$time.series[,1:3])

## Wood Duck

WODU.ts <- ts(WODU_ME_month$observation_per_min_avg, 
                             start = c(2010,1), frequency = 12)

WODU.ts.decomposed <- stl(WODU.ts, s.window = "periodic")
# Visualize the decomposed series. 
plot(WODU.ts.decomposed)

monthly_WODU_trend <- Kendall::SeasonalMannKendall(WODU.ts)
summary(monthly_WODU_trend)

WODU.ts.components <- as.data.frame(WODU.ts.decomposed$time.series[,1:3])

```











